Beyond clines: lineages and haplotype blocks in hybrid zones

### 2016-04-27 drafts/hybrid_zone.tex
submitting

The process of speciation commonly involves populations diverging with relatively little gene flow \citep{Coyne2004}.  
However, when formerly isolated populations come into contact before reproductive isolation is complete, some gene flow is possible.
Interbreeding and migration between such populations  
creates a gradient of alleles derived from the two populations across geographic space,
centered on their point of contact \citep[reviewed by][]{Barton1985}. 
If the populations are sufficiently diverged, 
this process leaves a distinct pattern of variation across the genome, 
in which long tracts of divergent haplotypes from each ancestral population  
are broken up by historical recombination events,
forming chromosomal {junctions} in ancestry \citep{Fisher1954, Chapman2002, baird2003distribution}. 
These patterns of correlated ancestry in admixed populations have been previously used to infer histories of hybridization and admixture 
\citep[e.g.,][]{Gravel2012,Hellenthal2014,Harris2013,sedghifar2015spatial}. 
Here we show how natural selection against hybrid incompatibilities shapes patterns of coancestry around such loci. %YB Comment. I really want people to know what we are doing by the end of the 1st paragraph

@@

Our work here builds on the large body of theory describing allele frequency clines in tension zones % YB added ' allele frequency clines'
and aims to describe \emph{transient} dynamics with a focus on \emph{haplotype} patterns.
% which may carry substantial information about the formation of recent zones.
To do this, we take the reverse-time perspective,
understanding the temporal dynamics of the system by describing how lineages tracing back from the hybrid zone
moveme across space, conditioning on the frequencies of the selected allele and following them as they recombine between selected backgrounds
\citep[similar to][]{hudson1988coalescent}.
Using a combination of theory and simulated hybrid zones, 
and extending the framework of \citet{sedghifar2015spatial},
we describe genome-wide patterns of coancestry 
as they relate to hybrid zone age, genetic distance from selected locus and geographic distance from hybrid zone center.

@@

To address this idea we develop analytical theory and a set of stochastic simulations. % YB Comment. i thought a paragraph starting with neglect was a downer
Our analytical theory provides a quantitative description of the problem, but in so doing, we neglect genetic drift  \revpoint{2}{9}
because of the well-known problems with spatial coalescence \citep{felsenstein1975pain,barton2002continuous}.
This should provide a good model for stochastic motion of a single lineage (our main focus),
but does not model correlations between lineages or non-Markovian dynamics of the lineages.
This approach is common to much theoretical work,
that studies the reaction-diffusion equations that govern the deterministic, high-density limit \citep[as in][]{Nagylaki1975},
or the discrete analogues \citep{hanson1966effects}. 
There is also substantial work on how clines are affected by genetic drift \citep{Slatkin1975,felsenstein1975genetic,Nagylaki1978,durrett2007width,barton2008effect,Polechova2011}.
Our simulation results  suggest that although genetic drift strongly affects local noisiness of the system,
it does not substantially influence our conclusions at the moderate population densities we examine,
perhaps because we study relatively short time-scales.
At much lower population densities, however, local extinction and recolonization may have a stronger effect. \linelabel{ll:much_lower_densities}

>>>>

The process of speciation commonly involves populations diverging with relatively little gene flow \citep{Coyne2004}.  
However, when formerly isolated populations come into contact before reproductive isolation is complete, some gene flow is possible.
Interbreeding and migration between such populations  
creates a gradient of alleles derived from the two populations across geographic space,
centered on their point of contact \citep[reviewed by][]{Barton1985}. 
If the populations are sufficiently diverged, 
this process leaves a distinct pattern of variation across the genome, 
in which long tracts of divergent haplotypes from each ancestral population  
are broken up by historical recombination events,
forming chromosomal {junctions} in ancestry \citep{Fisher1954, Chapman2002, baird2003distribution}. 
These patterns of correlated ancestry in admixed populations have been previously used to infer histories of hybridization and admixture 
\citep[e.g.,][]{Gravel2012,Hellenthal2014,Harris2013,sedghifar2015spatial}. 
Here we show how natural selection against hybrid incompatibilities shapes patterns of coancestry around such loci.

@@

Our work here builds on the large body of theory describing allele frequency clines in tension zones,
and aims to describe {transient} dynamics with a focus on {haplotype} patterns.
% which may carry substantial information about the formation of recent zones.
To do this, we take the reverse-time perspective,
understanding the temporal dynamics of the system by describing how lineages tracing back from the hybrid zone
moveme across space, conditioning on the frequencies of the selected allele and following them as they recombine between selected backgrounds
\citep[similar to][]{hudson1988coalescent}.
Using a combination of theory and simulated hybrid zones, 
and extending the framework of \citet{sedghifar2015spatial},
we describe genome-wide patterns of coancestry 
as they relate to hybrid zone age, genetic distance from selected locus and geographic distance from hybrid zone center.

@@

To address this idea we develop analytical theory and a set of stochastic simulations.
Our analytical theory provides a quantitative description of the problem, but in so doing, we neglect genetic drift \revpoint{2}{9}
(because of the well-known problems with spatial coalescence; \citet{felsenstein1975pain,barton2002continuous}).
This should provide a good model for stochastic motion of a single lineage (our main focus),
but does not model correlations between lineages or non-Markovian dynamics of the lineages.
This approach is common to much theoretical work,
that studies the reaction-diffusion equations that govern the deterministic, high-density limit \citep[as in][]{Nagylaki1975},
or the discrete analogues \citep{hanson1966effects}. 
There is also substantial work on how clines are affected by genetic drift \citep{Slatkin1975,felsenstein1975genetic,Nagylaki1978,durrett2007width,barton2008effect,Polechova2011}.
Our simulation results  suggest that although genetic drift strongly affects local noisiness of the system,
it does not substantially influence our conclusions at the moderate population densities we examine,
perhaps because we study relatively short time-scales.
At much lower population densities, however, local extinction and recolonization may have a stronger effect. \linelabel{ll:much_lower_densities}
